Supplemental Information: Using hydraulic geometry and the SWOT satellite to remotely sense river gas exchange velocity

### 2022-01-14 supp_v7.Rmd
some ms changes

---
title: "Supplemental Information: Exploiting hydraulic geometry to remotely sense gas exchange velocity via the SWOT satellite"
author:
  - Craig B Brinkerhoff:
      email: cbrinkerhoff@umass.edu
      institute: UMass
      correspondence: true
  - Peter A Raymond:
      institute: Yale
  - Colin J Gleason:
      institute: UMass
  - Christopher J Zappa:
      institute: Lamont
  - Merritt H Harlan:
      institute: UMass
institute:
  - UMass: Department of Civil & Environmental Engineering, University of Massachusetts, Amherst, MA
  - Yale: School of the Environment, Yale University, New Haven, CT
  - Lamont: Lamont-Doherty Earth Observatory, Columbia University, Palisades, NY
bibliography: References.bib
csl: agu_citation_style.csl
output:
  word_document:
      pandoc_args:
        - '--lua-filter=scholarly-metadata.lua'
        - '--lua-filter=author-info-blocks.lua'
      reference_docx: word-styles-reference.docx

@@

After joining this dataset with the hydraulics measurements provided with the 763 $k_{600}$ measurements, we classified each measurement as hydraulically wide if its hydraulic radius was within 4% of its mean flow depth. Hydraulic radius was calculated assuming a rectangular channel such that $R_H=2H+W$. Because these measurements were almost entirely made at streamgauges, which generally have a location bias favoring stable channels near bridges and other structures that yield lower width dynamics than those observed away from gauges [@allenPatternsRiverWidth2015; @parkWorldwideVariationsHydraulic1977], we used a wider threshold of 4%. In the 'width-limited' scenario common at streamgauges, the hydraulic radius is likely further from the mean flow depth than would occur naturally. The slightly more liberal threshold of 4% allowed for implicit accounting for this sampling bias.

## Text S2: Gas exchange model derivations
In this text we provide the full algebra to arrive at the four physically-based gas exchange models in the main text (equations 3-6). Equation 3 is derived via equations S1-S5. Equation 4 is derived via equations S6-S7. Equation 5 is derived via equations S8-S13. Equation 6 is derived via equations S14-S18. Consult Appendix A for all variable definitions. Note that $\alpha$ and $\beta$ are statistical parameters obtained via least squares regression for the small-eddy and chainsaw models, respectively. Subscripts denote variants of the same parameter, depending on which $\epsilon$ model is employed.

@@

**Chainsaw models**

@@

$A_0_i$ and $n_i$ prior hyperparameters were assigned following an updated version of the method developed by @brinkerhoffConstrainingRemoteRiver2020. They developed a set of river channel prior hyperparameters for McFLI algorithms that are entirely RS-able and reflect differential channel hydraulics as a function of river geomorphology. They used an extensive database of field measurements and machine learning to identify patterns that associate river width with the hydraulic priors needed to run McFLIs so that prior hyperparameters may be assigned to rivers using only the existing remotely sensed data. For this study, we extracted $A_0_{\lambda}$ and $A_0_{\gamma}$ as the 5th and 95th percentile values rather than the absolute maximum and minimum values to avoid physically impossible bounds on $A_0$.

This leaves the $k_{600_t}$ hyperparameters to be defined. $k_{600_{\mu}}$ is set by invoking the hydraulic geometry (HG) relationships developed in section 5.1.3 of the main text using the data from @brinkerhoffReconcilingAtaStationAtManyStations2019. We replaced both depth and velocity terms from our gas exchange model (equation 6 in the main text) with these HG models, resulting in equation S19 where Q is the mass-conserved streamflow for the river reach.

$$k_{600_{\mu}} = 62.82(gS_{i,t})^{7/16}(0.265Q_{t}^{0.395})^{9/16}*(0.284Q_{t}^{0.191})^{1/4} \mathbf{(S18)}$$

Obviously, we have no a priori information about Q. So, we use the Q prior that will be globally available when SWOT launches: a mean annual estimate from a water balance model. This is provided with the simulated SWOT data by both @durandIntercomparisonRemoteSensing2016 and @frassonExploringFactorsControlling2021. Again, using a temporally-invariant estimate of streamflow is the worst case scenario and BIKER's performance will improve with a more informed prior om streamflow (and therefore $k_{600}$). However as noted in the main text, our primary goal with this initial validation is to benchmark BIKER's worst case scenario for performance and so we do that here. Because BIKER treats every timestep as a different parameter, future work should investigate ways to assign temporally varying priors for $k_{600}$.

$k_{600_{\sigma}}$ is set to 0.30 (log-space). This corresponds to a coefficient of variation of approximately 0.30, which we took to reflect a reasonably strong agreement between the prior and the observed values. $k_{600_{\lambda}}$ was set to log(0.001) m/day. $k_{600_{\gamma}}$ was set to log(500) m/day. The

@@

Recall that $\sigma_{k_{600}}$ refers to the total uncertainty inherent in equation 8, i.e. stemming from both the chainsaw/form-drag dissipation model and the Manning's model for $\bar{U}$. For the purposes of this study, we are validating BIKER against equation 10 in the main text and so all uncertainties associated with the $\beta_1$ parameter are ignored and we only need to reflect the Manning's uncertainty in our specification of $\sigma_{k_{600}}^2$. Therefore, we take @hagemannBAMBayesianAMHGManning2017's estimated uncertainty from Manning's equation to infer streamflow from SWOT observations (0.25) and inflate it slightly to also account for the hydraulically-wide channel assumption and arrive at 0.30. This is the$\sigma_{k_{600}}$ used in this study.

In the scenario that BIKER is run on real SWOT data, $\sigma_k_{600}$ must reflect the full uncertainty implicit in equation 8 in the main text. This means we must also account for uncertainty from the $\beta_1$ parameter. Assuming perfect measurements from are made by the SWOT satellite, the $k_{600}$ relative uncertainty is expressed for some set of hydraulic observations as equation S19. We use the 166 hydraulically-wide measurements in our field-measured $k_{600}$ dataset and run 166 different Monte Carlo simulations (each 10,000 runs) to obtain 166 different $\sigma_k_{600}$ terms. We then take the average of those to be a reasonable estimate of $\sigma_k_{600}$. This ultimately provided a value of 1.1 for the log-transformed $\sigma_k_{600}$ that should be used once SWOT launches.

$$log_{k_{600}} \sim normal \left[ log(\beta_1), log(\sigma_{\beta_1}) \right] + \frac{1}{4}normal\left[log(\bar{U}), log(\sigma_{mannings})\right] \frac{7}{16}log(gS) + \frac{9}{16}log(H)  \mathbf{(S19)}$$

@@

|**Study**|**Workers**|**Number of measurements**|
|---------|-------------|--------------------------|
|@ulsethDistinctAirWater2019|@ulsethDistinctAirWater2019||
|@ulsethDistinctAirWater2019|@hallUseArgonMeasure2018||
|@ulsethDistinctAirWater2019|@schelkerCO2EvasionSteep2016||
|@ulsethDistinctAirWater2019|@mauriceInfluenceFlowBed2017||
|@ulsethDistinctAirWater2019|@raymondScalingGasTransfer2012||
|@churchillPredictionStreamReaeration1964|@churchillPredictionStreamReaeration1964||
|@owensReaerationStudiesStreams1964|@owensReaerationStudiesStreams1964||
<!-- |@raymondScalingGasTransfer2012|@melchingReaerationEquationsDerived| -->
<!-- |@raymondScalingGasTransfer2012|@tsivoglouCharacterizationStreamReaeration1972| -->
<!-- |@raymondScalingGasTransfer2012|Stroud Water Center| -->
<!-- |@raymondScalingGasTransfer2012|@bernotInterregionalComparisonLanduse2010| -->
<!-- |@raymondScalingGasTransfer2012|@mulhollandInterbiomeComparisonFactors2001| -->
<!-- |@churchillPredictionStreamReaeration1964|@churchillPredictionStreamReaeration1964|| -->
<!-- |@owensReaerationStudiesStreams1964|@owensReaerationStudiesStreams1964|| -->

>>>>

---
title: "Supplemental Information: Using hydraulic geometry and the SWOT satellite to remotely sense river gas exchange velocity"
author:
  - Craig B Brinkerhoff:
      email: cbrinkerhoff@umass.edu
      institute: UMass
      correspondence: true
  - Colin J Gleason:
      institute: UMass
  - Christopher J Zappa:
      institute: Lamont
  - Peter A Raymond:
      institute: Yale
  - Merritt H Harlan:
      institute: UMass
institute:
  - UMass: Department of Civil & Environmental Engineering, University of Massachusetts, Amherst, MA
  - Lamont: Lamont-Doherty Earth Observatory, Columbia University, Palisades, NY
  - Yale: School of the Environment, Yale University, New Haven, CT
bibliography: References.bib
csl: agu_citation_style.csl
output:
  word_document:
      pandoc_args:
        - '--lua-filter=scholarly-metadata.lua'
        - '--lua-filter=author-info-blocks.lua'
      reference_docx: word-styles-reference.docx

@@

Finally, we filtered this dataset to only measurements at least 100m wide ('SWOT-observable'). This left us with `r HG_swot$n` total river hydraulics measurements.

## Text S2: Gas exchange model derivations
In this text we provide the full algebra to arrive at the four physically-based gas exchange models in the main text (equations 3-6). Equation 3 is derived via equations S1-S5. Equation 4 is derived via equations S6-S7. Equation 5 is derived via equations S8-S13. Equation 6 is derived via equations S14-S18. Consult Appendix A for all variable definitions. Note that $\alpha$ and $\beta$ are statistical parameters obtained via least squares regression for the small-eddy and Reynolds models, respectively. Subscripts denote variants of the same parameter, depending on which $\epsilon$ model is employed.

@@

**Reynolds extension models**

@@

$A_0_i$ and $n_i$ prior hyperparameters were assigned following an updated version of the method developed by @brinkerhoffConstrainingRemoteRiver2020. They developed a set of river channel prior hyperparameters for McFLI algorithms that are entirely RS-able and reflect differential channel hydraulics as a function of river geomorphology. They used an extensive database of field measurements and machine learning to identify patterns that associate river width with the hydraulic priors needed to run McFLIs so that prior hyperparameters are assigned to rivers using only the remotely sensed measurements. For this study, we extracted $A_0_{\lambda}$ and $A_0_{\gamma}$ as the 5th and 95th percentile values rather than the absolute maximum and minimum values to avoid physically impossible bounds on $A_0$.

This leaves the $k_{600_t}$ hyperparameters to be defined. $k_{600_{\mu}}$ is set by invoking the hydraulic geometry (HG) relationships developed in section 3.2.3 of the main text using the data from @brinkerhoffReconcilingAtaStationAtManyStations2019. We replaced both depth and velocity terms from our gas exchange model (equation 6 in the main text) with these HG models, resulting in equation S19 where Q is the mass-conserved streamflow for the river reach.

$$k_{600_{\mu}} = 62.82(gS_{i,t})^{7/16}(0.265Q_{t}^{0.395})^{9/16}*(0.284Q_{t}^{0.191})^{1/4} \mathbf{(S19)}$$

Obviously, we have no a priori information about Q. So, we use the Q prior that will be globally available when SWOT launches: a mean annual estimate from a water balance model. This is provided with the simulated SWOT data by both @durandIntercomparisonRemoteSensing2016 and @frassonExploringFactorsControlling2021. Using a temporally-invariant estimate of streamflow is the worst case scenario and BIKER's performance will improve with a more informed prior on streamflow (and therefore $k_{600}$). This was recently shown on many of these rivers for SWOT discharge algorithms [@frassonExploringFactorsControlling2021]. However as noted in the main text, our primary goal with this initial validation is to benchmark BIKER's worst case scenario for performance and so we do that here.

Finally, $k_{600_{\sigma}}$ is set to 0.30 (log-space). This corresponds to a coefficient of variation of approximately 0.30, which we took to reflect a reasonably strong agreement between the prior and the observed values. $k_{600_{\lambda}}$ was set to log(0.001) m/day. $k_{600_{\gamma}}$ was set to log(500) m/day.

@@

Recall that $\sigma_{k_{600}}$ refers to the total uncertainty inherent in equation 8, i.e. stemming from both the Reynolds/form-drag dissipation model and the Manning's model for $\bar{U}$. For the purposes of this study, we are validating BIKER against equation 10 in the main text and so all uncertainties associated with the $\beta_1$ parameter are ignored and we only need to reflect the Manning's uncertainty in our specification of $\sigma_{k_{600}}^2$. Therefore, we take @hagemannBAMBayesianAMHGManning2017's estimated uncertainty from Manning's equation to infer streamflow from SWOT observations (0.25) and inflate it slightly to also account for the hydraulically-wide channel assumption and arrive at 0.30. This is the $\sigma_{k_{600}}$ used in this study.

In the scenario that BIKER is run on real SWOT data, $\sigma_k_{600}$ must reflect the full uncertainty implicit in equation 8 in the main text. This means we must also account for uncertainty from the $\beta_1$ parameter. Assuming perfect, no-error measurements are made by the SWOT satellite, the $k_{600}$ relative uncertainty is expressed for some set of hydraulic observations as equation S20. We use the 166 hydraulically-wide measurements in our field-measured $k_{600}$ dataset and run 166 different Monte Carlo simulations (each 10,000 runs) to obtain 166 different $\sigma_k_{600}$ terms. We then take the average of those to be a reasonable estimate of $\sigma_k_{600}$. This ultimately provided a value of 1.1 for the log-transformed $\sigma_k_{600}$ that should be used once SWOT launches.

$$log_{k_{600}} \sim normal \left[ log(\beta_1), log(\sigma_{\beta_1}) \right] + \frac{1}{4}normal\left[log(\bar{U}), log(\sigma_{mannings})\right] \frac{7}{16}log(gS) + \frac{9}{16}log(H)  \mathbf{(S20)}$$

@@

|**Study**|**Workers**|**Number of measurements**|
|---------|-------------|--------------------------|
|@ulsethDistinctAirWater2019|@ulsethDistinctAirWater2019|
|@ulsethDistinctAirWater2019|@hallUseArgonMeasure2018|
|@ulsethDistinctAirWater2019|@schelkerCO2EvasionSteep2016|
|@ulsethDistinctAirWater2019|@mauriceInfluenceFlowBed2017|
|@ulsethDistinctAirWater2019|@raymondScalingGasTransfer2012|
|@churchillPredictionStreamReaeration1964|@churchillPredictionStreamReaeration1964|
|@owensReaerationStudiesStreams1964|@owensReaerationStudiesStreams1964|
